---
title: "nichenet_endothelail"
output: html_document
date: "2024-03-25"
---

```{r}
library(Seurat)
library(nichenetr)
library(devtools)
library(cellranger)
library(dplyr)
library(ggplot2)
library(sctransform)
library(cowplot)
library(harmony)
library(tibble)
library(readxl)
library(Libra)
library(EnhancedVolcano)
library(BiocManager)
library(SeuratWrappers)
library(parallel)
library(SeuratDisk)
library(SeuratData)
library(monocle3)
library(patchwork)
library(ComplexHeatmap)
```

```{r}
library(Seurat)
devtools::install_github("saeyslab/nichenetr")
```

```{r}
library(nichenetr)
library(tidyverse)

#LS44_Feb24_subset <- UpdateSeuratObject(LS44_Feb24_subset)

```


```{r}
options(timeout=400)
lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
ligand_tf_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_tf_matrix.rds"))
```

```{r}
lr_network <- lr_network %>% distinct(from, to)

```

reciever
```{r}
receiver = c("Suprabasal Keratinocytes","Fibroblasts","Keratinocytes","Basal Keratinocytes","Granular Keratinocytes","Smooth Muscle Cells",        "Endothelial Cells","T Cells","Mast Cells", "Monocytes/Macrophages","Follicular Keratinocytes","Melanocytes/Neuronal Cells", "B Cells","Eccrine Glands")
receiver = c("Suprabasal Keratinocytes","Fibroblasts","Keratinocytes","Basal Keratinocytes","Granular Keratinocytes","Smooth Muscle Cells","Endothelial Cells","T Cells", "Monocytes/Macrophages","Follicular Keratinocytes", "B Cells")
expressed_genes_receiver <- get_expressed_genes(receiver, LS44_Feb24_subset, pct = 0.05)
```


Get a list of all receptors available in the ligand-receptor network, and define expressed receptors as genes that are in the ligand-receptor network and expressed in the receiver. Then, define the potential ligands as all ligands whose cognate receptors are expressed.
```{r}
all_receptors <- unique(lr_network$to)  
expressed_receptors <- intersect(all_receptors, expressed_genes_receiver)

potential_ligands <- lr_network %>% filter(to %in% expressed_receptors) %>% pull(from) %>% unique()
```

sender
```{r}
#unique(LS44_Feb24_subset@meta.data$celltype1)
sender_celltypes <-"Endothelial Cells"

# Use lapply to get the expressed genes of every sender cell type separately here
list_expressed_genes_sender <- sender_celltypes %>% unique() %>% lapply(get_expressed_genes, LS44_Feb24_subset, 0.05)
expressed_genes_sender <- list_expressed_genes_sender %>% unlist() %>% unique()

potential_ligands_focused <- intersect(potential_ligands, expressed_genes_sender) 
```

define gene set of interest 
```{r}
condition_oi <-  "LS"
condition_reference <- "Healthy"

seurat_obj_receiver <- subset(LS44_Feb24_subset, idents = receiver)

DE_table_receiver <-  FindMarkers(object = seurat_obj_receiver,
                                  ident.1 = condition_oi, ident.2 = condition_reference,
                                  group.by = "health",
                                  min.pct = 0.05) %>% rownames_to_column("gene")

geneset_oi <- DE_table_receiver %>% filter(p_val_adj <= 0.05 & abs(avg_log2FC) >= 0.25) %>% pull(gene)
geneset_oi <- geneset_oi %>% .[. %in% rownames(ligand_target_matrix)]
```

background expressed genes
```{r}
background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]

length(background_expressed_genes)
## [1] 5237
length(geneset_oi)
## [1] 2389
```

ligand activity analysis
```{r}
ligand_activities <- predict_ligand_activities(geneset = geneset_oi,
                                               background_expressed_genes = background_expressed_genes,
                                               ligand_target_matrix = ligand_target_matrix,
                                               potential_ligands = potential_ligands)

#ligand_activities <- ligand_activities %>% arrange(-aupr_corrected) %>% mutate(desc(aupr_corrected))
ligand_activities <- ligand_activities[order(-ligand_activities$aupr_corrected),]
ligand_activities
```
