---
title: "endothelial subset and nichenet"
output: html_document
date: "2024-03-28"
---
```{r}
library(Seurat)
library(nichenetr)
library(devtools)
library(cellranger)
library(dplyr)
library(ggplot2)
library(sctransform)
library(cowplot)
library(harmony)
library(tibble)
library(readxl)
library(Libra)
library(EnhancedVolcano)
library(BiocManager)
library(SeuratWrappers)
library(parallel)
library(SeuratDisk)
library(SeuratData)
library(monocle3)
library(patchwork)
library(ComplexHeatmap)
library(RColorBrewer)
library(tidyverse)
library(dittoSeq)

```
```{r}
endothelial_Mar24 <- Updateendothelial_Mar24ect(endothelial_Mar24)

DimPlot(LS44_Feb24_subset, label = T)
endothelial_Mar24@meta.data$library_id<- factor(x = endothelial_Mar24@meta.data$library_id, levels = c("SC124", "SC50", "SC68", "SC125", "SC1", "SC4", "SC18", "SC33", "SC392", "SC386", "SC407","SC32", "SC296", "SC297", "HSK053", "HSK054", "PHC003",  "SC275", "SC445", "SC446", "SC260", "SC222", "SC266", "SC267", "SC272", "HJ809", "HJ867", "SC389", "SC408", "SC409",     "SC126", "SC443", "SC198","SC246", "SC259", "SC144", "SC300", "SC388", "SC391", "SC424", "SC442", "SC457", "SC361", "SC455" )) # change the order of the factor levels

sample_h_info<- unique(data_frame(endothelial_Mar24@meta.data$library_id, endothelial_Mar24@meta.data$health, endothelial_Mar24@meta.data$onset))
sample_h_info
#sample_h_info
#order info: H adult, H ped, LS adult, LS ped
#H adult (8): "SC124", "SC50", "SC68", "SC125", "SC1", "SC4", "SC18", "SC33"
#H ped (9): "SC32", "SC296", "SC297", "HSK053", "HSK054", "PHC003","SC392", "SC386", "SC407"
#LS adult (13):  "SC275", "SC445", "SC446", "SC260", "SC222", "SC266", "SC267", "SC272", "HJ809", "HJ867", "SC389", "SC408", "SC409"
#LS ped (14): "SC126", "SC443", "SC198","SC246", "SC259", "SC144", "SC300", "SC388", "SC391", "SC424", "SC442", "SC457","SC455"
samples_ordered<- c("SC124", "SC50", "SC68", "SC125", "SC1", "SC4", "SC18", "SC33", "SC392", "SC386", "SC407","SC32", "SC296", "SC297", "HSK053", "HSK054", "PHC003",  "SC275", "SC445", "SC446", "SC260", "SC222", "SC266", "SC267", "SC272", "HJ809", "HJ867", "SC389", "SC408", "SC409",     "SC126", "SC443", "SC198","SC246", "SC259", "SC144", "SC300", "SC388", "SC391", "SC424", "SC442", "SC457","SC361" ,"SC455" )
```

```{r fig.height=6, fig.width=10}
FeaturePlot(endothelial_Mar24, features = c("IL33","IL6","NOTCH4","NOTCH1","JAG1","JAG2", "IL7"))

 FeaturePlot(endothelial_Mar24,split.by = "health", features = c("NOTCH4","NOTCH1","JAG1","JAG2"))
DotPlot(endothelial_Mar24, features = c("NOTCH4","NOTCH1","JAG1","JAG2") , split.by = "health")
install.packages("genemodulepheatmap")
library(pheatmap)
DoHeatmap(endothelial_Mar24, features = endo.marker.genes)
endo.marker.genes<- c("SEMA3G", "GJA4", "HEY1", "FBLN5", "ACKR1", "SOD2", "CLU", "VWF", "CCL14","CD36", "CA4", "FABP4", "SFN","DMKM","KLF5","PROX","LYVE", "SOX17", "PLAUR", "KLHL21", "GJA1", "LITAF", "ICAM1","EFNB2","RGS5")
FeaturePlot(endothelial_Mar24, features = "IL1R1", split.by = "health")
VlnPlot(endothelial_Mar24, group.by = "library_id", features = "IL33")+ FontSize(x.text=6)
DotPlot(endothelial_Mar24, group.by = "library_id", features = c("IL33", "IL1RL1", "IL1R4"))
DoHeatmap()
FeaturePlot(LS44_Feb24_subset, features = c("JAG1" , "JAG2", "NOTCH1", "NOTCH4"))
FeaturePlot(LS44_Feb24_subset, features = "NOTCH4")
unique(LS44_Feb24_subset@meta.data$library_id)
```



```{r}
options(timeout=400)
lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
ligand_tf_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_tf_matrix.rds"))
```

```{r}
lr_network = lr_network %>% distinct(from, to)
head(lr_network)
weighted_networks_lr = weighted_networks$lr_sig %>% inner_join(lr_network, by = c("from","to"))
head(weighted_networks$lr_sig)
```

Capillary EC            Post-capillary venules  Venous EC               Pre-venular Capillaries
[5] Pericytes/EC            Lymphatic EC            Proliferating EC        Arterial EC     

```{r}
## receiver
unique(Idents(endothelial_Mar24))
receiver = c("Suprabasal Keratinocytes",   "Fibroblasts", "Keratinocytes","Basal Keratinocytes",  "Granular Keratinocytes",    
 "Smooth Muscle Cells",   "Endothelial Cells","T Cells" ,  "Mast Cells","Monocytes/Macrophages","Follicular Keratinocytes",   "Melanocytes/Neuronal Cells", "B Cells", "Eccrine Glands" )
expressed_genes_receiver = get_expressed_genes(receiver, LS44_Feb24_subset, pct = 0.10)

background_expressed_genes = expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
FeaturePlot(endothelial_Mar24, features="IL6", split.by = "health")
## sender
sender_celltypes ="Endothelial Cells"
sender_celltypes=c("Capillary EC","Post-capillary venules",  "Venous EC","Pre-venular Capillaries","Pericytes/EC","Lymphatic EC",            "Proliferating EC","Arterial EC")
list_expressed_genes_sender = sender_celltypes %>% unique() %>% lapply(get_expressed_genes, LS44_Feb24_subset, 0.10) # lapply to get the expressed genes of every sender cell type separately here
expressed_genes_sender = list_expressed_genes_sender %>% unlist() %>% unique()
```
define gene set of interest
```{r}
seurat_obj_receiver= subset(LS44_Feb24_subset, idents = receiver)
seurat_obj_receiver = SetIdent(LS44_Feb24_subset, value = seurat_obj_receiver[["health", drop=TRUE]])

condition_oi = "LS"
condition_reference = "Healthy" 
  
DE_table_receiver = FindMarkers(object = seurat_obj_receiver, ident.1 = condition_oi, ident.2 = condition_reference, min.pct = 0.10) %>% rownames_to_column("gene")

geneset_oi = DE_table_receiver %>% filter(p_val_adj <= 0.05 & abs(avg_log2FC) >= 0.25) %>% pull(gene)
geneset_oi = geneset_oi %>% .[. %in% rownames(ligand_target_matrix)]
```

define ligands
```{r}
ligands = lr_network %>% pull(from) %>% unique()
receptors = lr_network %>% pull(to) %>% unique()

expressed_ligands = intersect(ligands,expressed_genes_sender)
expressed_receptors = intersect(receptors,expressed_genes_receiver)

potential_ligands = lr_network %>% filter(from %in% expressed_ligands & to %in% expressed_receptors) %>% pull(from) %>% unique()

```
ligand activity analysis
```{r}
ligand_activities = predict_ligand_activities(geneset = geneset_oi, background_expressed_genes = background_expressed_genes, ligand_target_matrix = ligand_target_matrix, potential_ligands = potential_ligands)

ligand_activities = ligand_activities %>% arrange(-aupr_corrected) %>% mutate(rank = rank(desc(aupr_corrected)))
ligand_activities
best_upstream_ligands = ligand_activities %>% top_n(30, aupr_corrected) %>% arrange(-aupr_corrected) %>% pull(test_ligand) %>% unique()

DotPlot(LS44_Feb24_subset, features = best_upstream_ligands %>% rev(), cols = "RdYlBu") + RotatedAxis()+FontSize(6)
```

```{r fig.width=10}
active_ligand_target_links_df = best_upstream_ligands %>% lapply(get_weighted_ligand_target_links,geneset = geneset_oi, ligand_target_matrix = ligand_target_matrix, n = 200) %>% bind_rows() %>% drop_na()
library(tidyverse)
active_ligand_target_links = prepare_ligand_target_visualization(ligand_target_df = active_ligand_target_links_df, ligand_target_matrix = ligand_target_matrix, cutoff = 0.33)

order_ligands = intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev() %>% make.names()
order_targets = active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links)) %>% make.names()
rownames(active_ligand_target_links) = rownames(active_ligand_target_links) %>% make.names() # make.names() for heatmap visualization of genes like H2-T23
colnames(active_ligand_target_links) = colnames(active_ligand_target_links) %>% make.names() # make.names() for heatmap visualization of genes like H2-T23

vis_ligand_target = active_ligand_target_links[order_targets,order_ligands] %>% t()
p_ligand_target_network = vis_ligand_target %>% make_heatmap_ggplot("Prioritized ligands","Predicted target genes", color = "purple",legend_position = "top", x_axis_position = "top",legend_title = "Regulatory potential")  + theme(axis.text.x = element_text(face = "italic")) + scale_fill_gradient2(low = "whitesmoke",  high = "purple", breaks = c(0,0.0045,0.0090))
p_ligand_target_network
```
receptors of top ranked ligands
```{r}
lr_network_top = lr_network %>% filter(from %in% best_upstream_ligands & to %in% expressed_receptors) %>% distinct(from,to)
best_upstream_receptors = lr_network_top %>% pull(to) %>% unique()

lr_network_top_df_large = weighted_networks_lr %>% filter(from %in% best_upstream_ligands & to %in% best_upstream_receptors)

lr_network_top_df = lr_network_top_df_large %>% spread("from","weight",fill = 0)
lr_network_top_matrix = lr_network_top_df %>% select(-to) %>% as.matrix() %>% magrittr::set_rownames(lr_network_top_df$to)

dist_receptors = dist(lr_network_top_matrix, method = "binary")
hclust_receptors = hclust(dist_receptors, method = "ward.D2")
order_receptors = hclust_receptors$labels[hclust_receptors$order]
    
dist_ligands = dist(lr_network_top_matrix %>% t(), method = "binary")
hclust_ligands = hclust(dist_ligands, method = "ward.D2")
order_ligands_receptor = hclust_ligands$labels[hclust_ligands$order]

order_receptors = order_receptors %>% intersect(rownames(lr_network_top_matrix))
order_ligands_receptor = order_ligands_receptor %>% intersect(colnames(lr_network_top_matrix))

vis_ligand_receptor_network = lr_network_top_matrix[order_receptors, order_ligands_receptor]
rownames(vis_ligand_receptor_network) = order_receptors %>% make.names()
colnames(vis_ligand_receptor_network) = order_ligands_receptor %>% make.names()
p_ligand_receptor_network = vis_ligand_receptor_network %>% t() %>% make_heatmap_ggplot("Ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential")
p_ligand_receptor_network
```
```{r}
# DE analysis for each sender cell type
# this uses a new nichenetr function - reinstall nichenetr if necessary!
DE_table_all = Idents(endothelial_Mar24) %>% levels() %>% intersect(sender_celltypes) %>% lapply(get_lfc_celltype, seurat_obj = endothelial_Mar24, condition_colname = "health", condition_oi = condition_oi, condition_reference = condition_reference, expression_pct = 0.10, celltype_col = NULL) %>% reduce(full_join) # use this if cell type labels are the identities of your Seurat object -- if not: indicate the celltype_col properly
DE_table_all[is.na(DE_table_all)] = 0

# Combine ligand activities with DE information
ligand_activities_de = ligand_activities %>% select(test_ligand, pearson) %>% rename(ligand = test_ligand) %>% left_join(DE_table_all %>% rename(ligand = gene))
ligand_activities_de[is.na(ligand_activities_de)] = 0

# make LFC heatmap
lfc_matrix = ligand_activities_de  %>% select(-ligand, -pearson) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities_de$ligand)
rownames(lfc_matrix) = rownames(lfc_matrix) %>% make.names()

order_ligands = order_ligands[order_ligands %in% rownames(lfc_matrix)]
vis_ligand_lfc = lfc_matrix[order_ligands,]

colnames(vis_ligand_lfc) = vis_ligand_lfc %>% colnames() %>% make.names()

p_ligand_lfc = vis_ligand_lfc %>% make_threecolor_heatmap_ggplot("Prioritized ligands","LFC in Sender", low_color = "midnightblue",mid_color = "white", mid = median(vis_ligand_lfc), high_color = "red",legend_position = "top", x_axis_position = "top", legend_title = "LFC") + theme(axis.text.y = element_text(face = "italic"))
p_ligand_lfc
```


```{r fig.width=10}
#Annotation
Idents(LS44_Feb24_subset) <- LS44_Feb24_subset@meta.data$RNA_snn_res.0.9
LS44_Feb24_subset@meta.data$library_id<- factor(x = LS44_Feb24_subset@meta.data$library_id, levels = c("SC18", "SC33", "HJ867", "SC390", "SC411"))
cluster_names <- read.csv("/ix1/ktorok/trh107/LS44_Feb24/overall_annotations.csv")
cluster_names <- dplyr::filter(cluster_names, !is.na(cluster))
new.cluster.ids <- cluster_names$celltype1
names(new.cluster.ids) <- levels(LS44_Feb24_subset)
new.cluster.ids
LS44_Feb24_subset <- RenameIdents(LS44_Feb24_subset, new.cluster.ids)
j$clustcelltype2 <- Idents(j)
DimPlot(LS44_Feb24_subset, label = T)
DimPlot(j)
DimPlot(LS44_Feb24_subset)
```
```{r fig.height=13, fig.width=9}
unique(Idents(LS44_Feb24_subset))
unique(Idents(endothelial_Mar24))
Idents(endothelial_Mar24)<- endothelial_Mar24@meta.data$
capec<- subset(endothelial_Mar24, idents = "Arterial EC")
subset()
my.deg.art<- FindMarkers(endothelial_Mar24, 
                            ident.1 = c("Arterial EC"), 
                            ident.2 = NULL)

read.csv()
  EnhancedVolcano(my.deg.art,
    lab = rownames(my.deg.art),
    x = 'avg_log2FC',
    y = 'p_val',
    title = 'Arterial EC versus ALL',
    FCcutoff = 0.5,
    pointSize = 2.0,
    labSize = 5.0)
```


 Capillary EC            Post-capillary venules  Venous EC               Pre-venular Capillaries Pericytes/EC           
[6] Lymphatic EC            Proliferating EC        Arterial EC 
DEG endo LS v healthy per subcluster
```{r fig.height=6, fig.width=5}
endothelial_Mar24$celltype.health <- paste(endothelial_Mar24$celltype2,endothelial_Mar24$health, sep = "_")
Idents(endothelial_Mar24) <- "celltype.health"
unique(endothelial_Mar24$celltype.health)

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Capillary EC_LS", ident.2 = "Capillary EC_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/CapillaryEC_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 ="Post-capillary venules_LS", ident.2 = "Post-capillary venules_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/postCapillaryvenules_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Venous EC_LS", ident.2 = "Venous EC_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/venousEC_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Pre-venular Capillaries_LS", ident.2 = "Pre-venular Capillaries_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/prevenCapillaryEC_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Pericytes/EC_LS", ident.2 = "Pericytes/EC_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/PericytesEC_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Lymphatic EC_LS", ident.2 = "Lymphatic EC_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/lymphEC_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Proliferating EC_LS", ident.2 = "Proliferating EC_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/prolifEC_lsvh.csv' )

endo.de <- FindMarkers(endothelial_Mar24, ident.1 = "Arterial EC_LS", ident.2 = "Arterial EC_Healthy", verbose = FALSE)
write.csv(endo.de,file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/ArterialEC_lsvh.csv' )


deg<-read.csv(file = '/ix1/ktorok/trh107/LS44_Feb24/endothelialDEGs/ArterialEC_lsvh.csv', )
EnhancedVolcano(deg,
    lab = deg$X,
    x = 'avg_log2FC',
    y = 'p_val',
    title = 'Arterial LS versus Healthy',
    FCcutoff = 0.5, pCutoff = 1,
    pointSize = 1.0,
    labSize = 3.0)
FeaturePlot(endothelial_Mar24, features = "JUN", split.by = "health")

```

```{r}
VlnPlot(object = endothelial_Mar24, features = "PECAM1", split.by = 'health', pt.size = .1, group.by = "celltype2")
Idents(endothelial_Mar24)<-endothelial_Mar24$celltype2
DotPlot(endothelial_Mar24,features = c("CCL21","HEY1","FBLN5","SOD2","CLU","ACKR1","CD36","CA4","FABP4","PROX1","LYVE1","SOX17","PLAUR","IRF1","SELE","RGS5","SFN","DMKM","KLF5"))
```




CELLCHAT
```{r}
devtools::install_github("jinworks/CellChat")
install.packages('NMF')
library(NMF)
library(CellChat)
devtools::install_github("jokergoo/circlize")

```

```{r}
unique(Idents(LS44_Feb24_subset))
cellchat <- createCellChat(object = endothelial_Mar24, group.by = "celltype2", assay = "RNA")
CellChatDB <- CellChatDB.human
```
```{r}
# subset the expression data of signaling genes for saving computation cost
cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
#> The number of highly variable ligand-receptor pairs used for signaling inference is 692

execution.time = Sys.time() - ptm
print(as.numeric(execution.time, units = "secs"))
#> [1] 13.20763
# project gene expression data onto PPI (Optional: when running it, USER should set `raw.use = FALSE` in the function `computeCommunProb()` in order to use the projected data)
# cellchat <- projectData(cellchat, PPI.human)
```
 [1] Suprabasal Keratinocytes   Fibroblasts                Keratinocytes              Basal Keratinocytes        Granular Keratinocytes    
 [6] Smooth Muscle Cells        Endothelial Cells          T Cells                    Mast Cells                 Monocytes/Macrophages     
[11] Follicular Keratinocytes   Melanocytes/Neuronal Cells B Cells                    Eccrine Glands 

CIRCOS VIZ
```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```
